Staged Abstract Interpreters

### 2019-03-24 paper/20_prelim.tex
sec 2

\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursions
  case class AOp(op: String, e1: Expr, e2: Expr) extends Expr // primitives
\end{lstlisting}

@@

\subsection{Monads in Scala}

A monad is a type constructor @M[_]: * -> *@ with two operations, often called @return@ and @bind@.
Informally, @return@ wraps a value into the monad @M@, and @bind@ unwraps the monadic value and 
transforms it to a new monadic value.
Pragmatically in Scala, we define a monad using trait @Monad@ (Figure \ref{fig:monad}),
where it declares the @pure@ \footnote{We elect to use \texttt{pure} as the name, 
since \texttt{return} is a keyword in Scala and \texttt{unit} is a built-in function in LMS.}
and @flatMap@ operation. The trait itself takes the monad type @M[_]@ as
argument, which is a higher-kinded type that takes a type and returns a type. 
The method @pure@ promotes values of type @A@ to values of type @M[A]@.
The monadic @bind@ operation is
usually called @flatMap@ in Scala, which takes a monad-encapsulated value of
type @M[A]@, a function of @A => M[B]@ and returns values of type @M[B]@.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {                                  
    def pure[A](a: A): M[A]                            
    def flatMap[A,B](fa: M[A])(f: A => M[B]): M[B]     
  }                                                    
    \end{lstlisting}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}

@@

\todo{}

Before showing the actual implementation of the concrete semantics, let us first
consider a generic interface for the interpreters, which is semantics-agnostic
and can be "interpreted" by either concrete or abstract, unstaged or staged
semantics of our choice. The trait \texttt{Semantics} first declares several
abstract type members, such as \texttt{Addr}, \texttt{Value}, \texttt{Env},
\texttt{Store} and returned type \texttt{Ans}. It also declares initial values
for environments and stores, and few abstract methods that manipulate
environments and stores such as @put@ and @get@.

\begin{lstlisting}
trait Semantics {
  type R[+_]
  type Ident = String;  type Addr;  type Value;  type Env
  type Store;  type Ans = (R[Value], R[Store])
  def get($\rho$: R[Env], x: Ident): R[Addr]
  def put($\rho$: R[Env], x: Ident, a: R[Addr]): R[Env]
  def get($\sigma$: R[Store], a: R[Addr]): R[Value]
  def put($\sigma$: R[Store], a: R[Addr], v: R[Value]): R[Store]
  def alloc($\sigma$: R[Store], x: Ident): R[Addr]
  val $\rho$0: R[Env]; val $\sigma$0: R[Store]
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: R[Env]): R[Value]
  def num(i: Lit): R[Value]
  def apply_closure(ev: EvalFun)
    (f: R[Value], arg: R[Value], $\sigma$: R[Store]): Ans
  def branch0(cnd: R[Value], thn: => Ans, els: => Ans): Ans
  def prim_eval(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
}  // to be continued
\end{lstlisting}

The actual semantics and operations of the interpreter are left to be
implemented later, these include operations such as lifting literal terms to
values (\texttt{close}, \texttt{num}), applying a closure
(\texttt{apply\_closure}), branching (\texttt{branch0}), and arithmetics
(\texttt{prim\_eval}). We can choose to implement them in a way that is concrete
or abstract, unstaged or staged. But, just based on these abstract operations,
an expressive generic interpreter can be defined:

\begin{lstlisting}
trait Semantics {  // continued
  type EvalFun = (Expr, R[Env], R[Store]) => Ans
  def eval(ev: EvalFun)(e: Expr, $\rho$: R[Env], $\sigma$: R[Store]): Ans = 
    e match {
      case Lit(i) => (num(Lit(i)), $\sigma$)
      case Var(x) => (get($\sigma$, get($\rho$, x)), $\sigma$)
      case Lam(x, e) => (close(ev)(Lam(x, e), $\rho$), $\sigma$)
      case App(e1, e2) =>
        val (e1v, e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        apply_closure(ev)(e1v, e2v, e2$\sigma$)
      case Rec(x, f, body) =>
        val $\alpha$ = alloc($\sigma$, x)
        val $\rho$_* = put($\rho$, x, $\alpha$)
        val (fv, f$\sigma$) = ev(f, $\rho$_*, $\sigma$)
        val $\sigma$_* = put(f$\sigma$, $\alpha$, fv)
        ev(body, $\rho$_*, $\sigma$_*)
      case If0(cnd, thn, els) =>
        val (cndv, cnd$\sigma$) = ev(cnd, $\rho$, $\sigma$)
        branch0(cndv, ev(thn, $\rho$, cnd$\sigma$), ev(els, $\rho$, cnd$\sigma$))
      case AOp(op, e1, e2) =>
        val (e1v, e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        (prim_eval(op, e1v, e2v), e2$\sigma$)
    }
  def fix(ev: EvalFun => EvalFun): EvalFun = 
    (e, $\rho$, $\sigma$) => ev(fix(ev))(e, $\rho$, $\sigma$)
  def eval_top(e: Expr): Ans = eval_top(e, $\rho$0, $\sigma$0)
  def eval_top(e: Expr, $\rho$: R[Env], $\sigma$: R[Store]): Ans = 
    fix(eval)(e, $\rho$, $\sigma$)
}
\end{lstlisting}

The implementation of the skeleton interpreter traverses the abstract syntax
tree and should be straightforward to understand. It is worth noting that the
interpreter is written in open-recursive style -- it can not refer to itself
directly, instead, \texttt{eval} takes an additional parameter \texttt{ev}
referring to itself; this allows us to instrument the interpreter from the
outside. Accordingly, the function \texttt{close} for lifting lambda term to
closure values is also written in this style. We define a default implementation
\texttt{fix} to close the recursion by recursively using the result of
\texttt{fix(ev)} for \texttt{ev}. Finally, the top-level evaluation function
\texttt{eval\_top} is defined.

\paragraph{Binding-time Abstraction}
%\paragraph{Stage Polymorphism}
% Binding-time abstraction via abstract types

Readers may notice that an abstract higher-kinded type \texttt{R[+T]} is defined
in the trait, i.e. a type-level function which will be used to specify the
binding-time, i.e., staged or not; the annotation \texttt{+} indicates that the
type parameter \texttt{T} is covariant. The LMS framework we use here allows us
to specify binding-times purely based on types. All occurrences of type
\texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} essentially
because they may potentially be dynamic when staging is introduced. Obviously,
the unstaged higher-kinded type then can be simply instantiated as an identity
type function: \texttt{type R[+T] = T}. In Section~\ref{stagedinterp},
\texttt{R} will be instantiated as a staging annotation.

\todo{Finally tagless, poly embedding}

\subsection{Concrete Instantiation} \label{unstaged_conc}

>>>>

\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursions
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetics
\end{lstlisting}

@@

\subsection{Monads in Scala} \label{monadscala}

A monad is a type constructor @M[_]: * -> *@ with two operations, often called
@return@ and @bind@. Informally, @return@ wraps a value into the monad @M@, and
@bind@ unwraps the monadic value and transforms it to a new monadic value.
Pragmatically in Scala, we define a monad type class using trait @Monad@ (Figure
\ref{fig:monad}), where it declares the @pure@ \footnote{We elect to use
\texttt{pure} as the name, since \texttt{return} is a keyword in Scala and
\texttt{unit} is a built-in function in LMS.} and @flatMap@ operation. The trait
itself takes the monad type @M[_]@ as argument, which is a higher-kinded type
that takes a type and returns a type. The method @pure@ promotes values of type
@A@ to values of type @M[A]@. The monadic @bind@ operation is usually called
@flatMap@ in Scala, which takes a monad-encapsulated value of type @M[A]@, a
function of @A => M[B]@ and returns values of type @M[B]@.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {                                  
    def pure[A](a: A): M[A]                            
    def flatMap[A,B](ma: M[A])(f: A => M[B]): M[B]     
  }                                                    
    \end{lstlisting}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}

@@

\paragraph{Basic type definitions} We start with some basic type definitions.
The identifiers in the program are represented by strings. The two default
components in the interpreter are environments @Env@ and stores @Store@, i.e.,
mappings from identifiers to addressess and mappings from addresses to values,
respectively. But the @Addr@ and @Value@ are just declared as abstracrt types.

\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

\paragraph{Binding-time abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. If we instantiate @R@ as an identity
type (i.e., @type R[T] = T@), then the generic interpreter will execute the program.
In Section \ref{stagedinterp}, we will instantiate @R@ using LMS's built-in
next-stage type annotation @Rep@, which makes the interpreter act as a compiler.

\paragraph{Moandic operations} We define the return type of the interpreter as
@Ans@, which is a monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain constraints have to be posed onto the type @AnsM@. Here, we use a
structural type @MonadOps@ to require @AnsM@ to at least implement @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time; accordingly the outer @R@ defined in the trait is passed
to @MonadOps@. Inside of @MonadOps@, @R[_]@ wraps the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When acting as
compilers, we will also replace the monads to the ones that work on staged
values.

We also defines several methods to obtain and modify the environment and store.
These methods either returns monadic values with environment or store inside, or
simply a @Unit@ value for effects. For example, @ext_env@ takes a value @ans@ of
type @Ans@ and a pair of binding that maps @x@ to $\alpha$, and installs this
binding to the environment when evaluating @ans@.

\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \begin{lstlisting}
  type MonadOps[R[_], M[_], A] = {
    def map[B](f: R[A] => R[B]): M[B]
    def flatMap[B](f: R[A] => M[B]): M[B]
  }

  type AnsM[T] <: MonadOps[R, AnsM, T]
  type Ans = AnsM[Value]
    \end{lstlisting}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
// Environment operations
def ask_env: AnsM[Env]
def ext_env(ans: Ans)(xα: (Ident, R[Addr])): Ans
// Store operations
def get_store: AnsM[Store]
def put_store(σ: R[Store]): AnsM[Unit]
def set_store(av: (R[Addr], R[Value])): AnsM[Unit]
    \end{lstlisting}
  \end{subfigure}
\end{figure}
\vspace{-1em}

\paragraph{Primitive operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier, and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is also wrapped
by @R[_]@. The other one would simply wrap the address with our monadic type
@AnsM[_]@.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the interpreter.
The method @num@ and @close@ deal with primitive values, which lift literal
terms (e.g., lambdas) to our value representation (e.g., closures).
The method @get@ simply retrieves the value mapped from the identifier @x@ in
the environment and store. Conditionals and arithmetics is handled by @br0@
and @arith@, respectively. The methods @ap_clo@ takes a function value and an
arugment value and then does the application.

%TODO: close/ap_clo takes an ev

\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(fun: R[Value], arg: R[Value]): Ans
\end{lstlisting}

\paragraph{The interpreter} Now we can define the semantics-agnostic interpreter
in monadic form. The essential idea is to traverse the abstract syntax tree
while maintaining the effects such as reader and state. 
It is worth noting that the interpreter is written in open-recursive style -- it
can not refer to itself directly, instead, @eval@ takes an additional parameter
@ev@ of type @Expr => Ans@ to refer to itself. Accordingly, the method @close@
that lifts lambda terms to closures and @ap_clo@ that applies functions also
takes an extra @ev@.

\begin{lstlisting}
  def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
    case Lit(i) => num(i)                      case Let(x, rhs, e) => for {
    case Var(x) => for {                         v <- ev(rhs)
      ρ <- ask_env                               α <- alloc(x)
      σ <- get_store                             _ <- set_store(α → v)
     } yield get(σ, ρ, x)                        rt <- ext_env(ev(e))(x → α)
    case Lam(x, e) => for {                    } yield rt
      ρ <- ask_env                             case Aop(op, e1, e2) => for {
    } yield close(ev)(Lam(x, e), ρ)              v1 <- ev(e1)                                               
    case App(e1, e2) => for {                    v2 <- ev(e2)
      v1 <- ev(e1)                             } yield arith(op, v1, v2)
      v2 <- ev(e2)                             case Rec(x, rhs, e) => for {
      rt <- ap_clo(ev)(v1, v2)                   α <- alloc(x)
    } yield rt                                   v <- ext_env(ev(rhs))(x → α)
    case If0(e1, e2, e3) => for {                _ <- set_store(α → v)
      cnd <- ev(e1)                              rt <- ext_env(ev(e))(x → α)
      rt <- br0(cnd, ev(e2), ev(e3))           } yield rt                    
    } yield rt
  }
\end{lstlisting}

Since the interpreter is written in open-recursive style, we declare an abstract
combinator @fix@ used to close the recursion. For concrete-interpretation instantiation, it
works like the Y combinator; for abstract-interpretation instantiation, we will
instrument the interpreter by defining a memorized version of @fix@ to ensure
termination. Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, thus is also an abstract type.

\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

%==========================================================================

\subsection{A Concrete Interpreter} \label{unstaged_conc}
